# CHIRPS data download and basic manipulation reference guide

## Elaborated by: Juan Miguel Víquez, Daniel Knopp & Jorge Masis
### 1. Install the required packages
```{r}
	#install.packages("terra")
	#install.packages("rasterVis")
```
Terra is used for methods for spatial data analysis with raster and vector data.
More info on Terra: 
https://cran.r-project.org/web/packages/terra/index.html

### 2. Load the packages in the Rstudio session
```{r}
	library(rasterVis)
	library(terra)
```

### 3. Create the path for the Chirps dataset
This is the directory in the PC where Chirps is stored, change it to your computer's directory, accordingly. 
In my case is:
```{r}
	path <- "E:/Thao/chirps_monthly"
```
### 4. Read the names of the files and create a list() from them
```{r}	
	Chirps_files <- list.files(path,
                           pattern = NULL,
                           all.files = FALSE,
                           full.names = TRUE)
head(Chirps_files)
tail(Chirps_files)
```
### 4. Create a stack 
More info: https://r-spatial.github.io/stars/
```{r}
    chirps <- rast(Chirps_files)
chirps
```
A raster stack has 3 dimensions (X, Y, Time), each of the rasters have the monthly average precipitation and each of the pixels in these rasters contains the precipitation value on the location for that month.
If we want to plot the data, since the stack has 3 dimensions we use variablename[[]] to access a specific time:
```{r}
    plot(chirps[[1]])
```
Given that CHIRPS is a global product we need to crop the area to our study region. In order to do this we use a Shapefile as a "cookie cutter" to extract only the within our area of interest.
To load a shapefile we use the *vect* function of the terra package **(*Be aware of only using the .shp file and not the others ending on .shx or any other extension)***:
```{r}
    Aguan_basin_path <- "E:/Basin_Rapel/Aguan/Cuenca_Rio_Aguan.shp"
    Aguan_basin <- vect(Aguan_basin_path)
    plot(Aguan_basin)
```
### Adjusting the reference system of the Spatvector
####if the crs of the stack and the SHP does not match, we need to reproject the shapefile to the coordinates system of the raster, we can check that with 
```{r}
chirps
```


```{r}
Aguan_basin
```
####If the stack and the SHP are in a diferent coordinate system, we solve this with the project() function
####More info on https://rdrr.io/cran/terra/man/project.html
```{r}
  Aguan_basin<-project(Aguan_basin,"+proj=longlat +datum=WGS84")
Aguan_basin
```


### 5. We can start to cut our study area
For this, we need to use 2 functions. The first one is *crop()* (more info on https://rdrr.io/cran/terra/man/crop.html)
```{r}
    chirps_crop <- crop(chirps, Aguan_basin, snap = 'out')
chirps_crop
```
Plot the data
```{r}
    plot(chirps_crop[[1]])

##Show the outline of the basin

    lines(Aguan_basin)
```
The second function is *mask()*, this one eliminates any NA's resulting from the crop.
***(Always crop first and mask second, otherwise you will try to mask the whole extend of the datasheet and it takes a lot of processing and thus time)***
```{r}
    chirps_aguan <- mask(chirps_crop, Aguan_basin)
    plot(chirps_aguan[[1]])
    lines(Aguan_basin)
    
```
    
### 6. Get a monthly average plot and monthly string 
With *tapp()* we will grab the cropped raster stack and average for the index time period, then in order to get the monthly average we need to use an *index=1:12*, with that done we can assign the names of the months to the raster layer 
```{r}
    monthly_chirps<-tapp(chirps_aguan, index = 1:12, fun = mean)
    names(monthly_chirps)<-c('Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec')
    monthly_chirps

```

Next, we are make a *levelplot* for the monthly averages (see https://www.rdocumentation.org/packages/lattice/versions/0.10-10/topics/levelplot) 
```{r}
    levelplot( monthly_chirps,layout=c(4,3),contour=T,col.regions=hcl.colors(100))
```
### 7. Extract mean precipitation
This takes the already cropped and masked stack for extracting the values inside a specific shapefile.
```{r}
    extracted_values <- extract(chirps_aguan,Aguan_basin, fun = "mean")
head(extracted_values)
```
We transform the list into a *dataframe* and then transpose, this way we can get a *date/value* format
The first line is avoided, as it corresponds to the rows name:
```{r}
extracted_values <- as.numeric(extracted_values[2:481])
values <- as.data.frame(extracted_values)
head(values)
```
Following this, we can create a vector from the starting month to the last one. Using the file *basename* (i.e. "chirps-v2.0.2020.09.tif") we select from 2020 to 09 and use that to create a vector
```{r}
    Dates <- substr(basename(Chirps_files), 13, 19)
head(Dates)
```
The next step is to attach the dates and the *extracted_values* in a table 
```{r}
    Precipitation_Aguan <- cbind(Dates, values)
head(Precipitation_Aguan) 

```
### 7. Artifical stations using Chirps
Another method is creating a point layer and extract the locations from there
First we are creating a table of X (lon) and Y (lat) coordinates, then with the function vect() we create the point vector, for changing the coordinates make sure the coordinate system of your reference is also on WGS84 otherwise we will run with the same issue as the shapefile from the crop 
```{r}
xy <- cbind(c(-87.000,-86.500,-86.000),c(15.202,15.403,15.800))
p <- vect(xy, crs="+proj=longlat +datum=WGS84")
plot(Aguan_basin)
points(p)
```
###9 Extract several point from a raster stack 
Now that we got the coordinates of our points of interest we can extract the values for the point layer. 
#Note:The extract function can also use other functions (max,min,sum)
```{r}
extracted_values_points <- extract(chirps_Aguan, p, fun = "mean")
```

We need to crop the index row so we eliminate the first one from the data frame so we are gonna follow the same procedure 
```{r}
values_points <- (as.data.frame(extracted_values_points[2:481]))
Dates <- substr(basename(Chirps_files), 13, 19)
head(values_points)
```
###9: Creating a data frame for the extracted values 
Now we transpose the rows with the columns to create a table and then bind the rows with the Dates vector   
Precipitation_Rapel<-t(rbind(Dates,values_points))
Precipitation_Rapel<-as.data.frame(Precipitation_Rapel)
##Next we assign the headers for our table 
colnames(Precipitation_Rapel)<-c('date','lower','middle','upper')
##and saving it as a CSV file with write.csv()
##For saving files please remember to use the file extension of the format you want to save in 
##For this case a CSV
write.csv(Precipitation_Rapel,"E:/Basin_Rapel/Cuenca_Rapel4.csv")



```{r}

```

